% Document class for LME theses: lmedoc
% Master's Thesis (English)
\documentclass[english,mt]{lmedoc/lmedoc}

%%%%%%%%%%%%%%%%%%
% pdflatex and lualatex are supported
% ++ "Umlaut" support
\usepackage{iftex}
\ifPDFTeX
  \usepackage[utf8]{inputenc}
  \usepackage[T1]{fontenc}
  \usepackage{lmodern}
\else
  \ifXeTeX
     \usepackage{fontspec}
  \else 
     \usepackage{luatextra}
  \fi
\fi

% Load additional packages
\usepackage{lmedoc/packages}
% Additional packages for this thesis
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amsmath}
\usepackage{amssymb}
% subcaption already loaded by lmedoc/packages.sty
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{xcolor}
\usepackage{float}
\usepackage{tikz}
\usetikzlibrary{patterns}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

% put in this file your abbreviations
\input{abbreviations.tex}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% supress messages because of underful hbox/vbox, this is not a problem
\hbadness=10000
\vbadness=10000
\raggedbottom

% some useful commands
\makeatletter % let's define a single dot
\DeclareRobustCommand\onedot{\futurelet\@let@token\@onedot}
\newcommand{\@onedot}{\ifx\@let@token.\else.\null\fi\xspace}
\makeatother

% Sets the bib file
\addbibresource{literature.bib}

\pagenumbering{roman}

\begin{document}

% Cover page for Master's thesis
\begin{deckblatt}
  \Titel{Cold Diffusion for CT Field-of-View Extension} % Title
  \Name{Qianxin} % Last name
  \Vorname{Wang} % Given name
  \Geburtsort{Anhui} % Place of birth
  \Geburtsdatum{02.05.1998} % Date of birth
  \Betreuer{Prof. Dr.-Ing. Andreas Maier, Yipeng Sun} % Advisor
  \Start{12.09.2025} % Start of thesis
  \Ende{12.03.2026} % End of thesis
  %\ZweitInstitut{} % Optional: Cooperation partner
\end{deckblatt}

\cleardoublepage

% Declaration (required for thesis)
I hereby declare that this thesis is entirely my own work and that I have used
no sources or aids other than those stated. All passages that have been taken
verbatim or essentially from published or unpublished sources have been marked
as such. The thesis has not been submitted, either in the same or a similar form,
to any other examination board.

\vspace{2cm}

\noindent
\rule{0.45\textwidth}{0.4pt} \hfill \rule{0.45\textwidth}{0.4pt}\\
\noindent
\textit{Place, Date} \hfill \textit{Signature}

\cleardoublepage

% Abstract in English
\begin{center}
\bfseries
{\selectlanguage{english}Abstract}
\normalfont
\end{center}

\vspace{0.5cm}

Computed Tomography (CT) truncation artifacts occur when parts of the anatomy extend
beyond the scanner field-of-view, producing incomplete projection data (sinograms) and
artifacted reconstructions. This thesis investigates CT field-of-view extension in the
projection domain and treats the problem as recovery of missing sinogram regions under
controlled bilateral truncation. The proposed method adapts \textit{Cold Diffusion} to
this setting by replacing stochastic noising with deterministic masking that follows the
truncation process~\cite{bansal2022cold}. A time-conditioned U-Net predicts clean
sinograms from progressively truncated inputs, and reverse sampling is performed with
mask-guided updates together with hard replacement in observed regions to preserve
measured projection values.

Experiments are conducted on two public CT datasets, CT-ORG and LIDC-IDRI, to include
both abdominal and thoracic anatomy~\cite{rister2020ctorg,armato2011lung}. The study
uses a unified preprocessing and evaluation protocol and reports both sinogram-domain and
reconstruction-domain metrics. Baseline comparisons include truncated input, WCE-only
extrapolation, U-Net inpainting, conditional DDPM, and Cold Diffusion under matched
conditions. In the main baseline setting at \(k_c=320\) (50\% of detector columns truncated), U-Net inpainting achieves the
best quantitative performance, while Cold Diffusion remains consistently stronger than
conditional DDPM. A dedicated timestep study further shows that reverse-step budget
substantially influences reconstruction quality.

The results indicate that Cold Diffusion is a feasible projection-domain approach for CT
field-of-view extension and provides meaningful improvements over the Gaussian-noise
diffusion baseline in this setup. At the same time, performance remains sensitive to
sampling configuration and does not surpass the strongest single-step baseline under the
current protocol. These findings position the work as a practical and evidence-based
step toward diffusion-based projection restoration, with further validation needed for
broader truncation patterns and clinical deployment scenarios.

\textbf{Keywords:} Computed Tomography, Field-of-View Extension, Truncation Artifacts,
Cold Diffusion, Medical Image Processing, Sinogram Restoration

\cleardoublepage

\tableofcontents

\cleardoublepage 
\pagenumbering{arabic}

\include{introduction}
\cleardoublepage
\include{background}
\cleardoublepage
\include{related_work}
\cleardoublepage
\include{methodology}
\cleardoublepage
\include{experiments}
\cleardoublepage
\include{results}
\cleardoublepage
\include{discussion}
\cleardoublepage
\include{conclusion}
\cleardoublepage

\appendix
\cleardoublepage
\include{appendix_a_data_processing}
\cleardoublepage

%% Auto-generated lists %%
% Glossar
\printunsrtglossary[type=abbreviations]
\cleardoublepage

% List of figures
\addcontentsline{toc}{chapter}{\listfigurename}
\listoffigures
\cleardoublepage

% List of tables
\addcontentsline{toc}{chapter}{\listtablename}
\listoftables
\cleardoublepage

% Literature list
\selectlanguage{english}{\addcontentsline{toc}{chapter}{\bibname}}
\printbibliography

\end{document}
